---
title: "Analysis of the Properties Influencing the Quality of Red Wine"
author: "Grayson Ricketts"
date: "March 21, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load the data and install packages, include = FALSE}
rw <- read.csv('https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv')

library(ggplot2)
library(dplyr)
library(scales)
library(gridExtra)
```

***

### Guiding question: 
Which chemical properties influence the quality of red wines?

***

### Basic data exploration and Summary

```{r Data structure and size, echo = FALSE}
dim(rw)
names(rw)
str(rw)
```
[data set description](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt)  

```{r Quality graphs, echo = FALSE}
ggplot(data = rw, aes(quality)) +
  geom_histogram(binwidth = 1, color = 'red') +
  scale_x_continuous(breaks=seq(1,10, 1)) +
  ggtitle("Quality (Ranking)") +
  labs(y = 'count')

ggplot(data = rw, aes(quality)) +
  geom_histogram(binwidth = 1, color = 'red') +
  scale_x_continuous(breaks=seq(1,10, 1)) +
  scale_y_log10(breaks=seq(0, 1000, 250)) +
  ggtitle("Quality (Ranking)") +
  labs(y = 'count (log10)')

table(rw$quality)
```
The vast majority of the wines got ranked 5 and 6 with each rank away from 5 and 6 having a magnitude less number

```{r ph, Alcohol, Fixed Acidity univariate graphs, echo = FALSE}
ggplot(data = rw, aes(pH)) +
  geom_histogram(binwidth = .05, color = 'red') +
  scale_x_continuous(breaks=seq(1,5, .1)) +
  ggtitle("pH")

ggplot(data = rw, aes(alcohol)) +
  geom_histogram(binwidth = .5, color = 'red') +
  scale_x_continuous(breaks=seq(0,20, .5)) +
  ggtitle("Alcohol")

ggplot(data = rw, aes(fixed.acidity)) +
  geom_histogram(binwidth = .2, color = 'red') +
  scale_x_continuous(breaks=seq(0,20, 1)) +
  ggtitle("Fixed Acidity")
```
Overall acidity, alcohol level, and the fixed acidity all are normally distributed with positive skewness on acohol and fixed acidity.

```{r Citric acid, echo=FALSE}
ggplot(data = rw, aes(citric.acid)) +
  geom_histogram(binwidth = .01, color = 'red') +
  scale_x_continuous(breaks=seq(0,1, .1)) +
  ggtitle("Citric Acid")
```
Citric acid in a large number of wines is 0, and the distribution is positively--and noticeable flat-- skewed.  

#### Acid summaries (Fixed, volatile, citric)
```{r All acids, echo=FALSE}
summary(rw$fixed.acidity)
summary(rw$volatile.acidity)
summary(rw$citric.acid)
```
Each acidity measurement seems to be different. Fixed acidity seems to be positively skewed while volatile acidity is less harshly skewed but has some positive outliers while citric acid has a almost exponential distribution.

```{r Chlorides histogram, echo=FALSE}
ggplot(data = rw, aes(chlorides)) +
  geom_histogram(binwidth = 0.025, color = 'red') +
  scale_x_continuous(breaks = seq(0, 1, .05))

ggplot(data = rw, aes(chlorides)) +
  geom_histogram(binwidth = 0.05, color = 'red') +
  scale_x_log10(breaks = pretty_breaks(rw$chlorides, n = 10)) +
  labs(x = 'chlorides (log10)')

ggplot(data = subset(rw, chlorides < quantile(chlorides, .98)), 
       aes(chlorides)) +
  geom_histogram(binwidth = 0.005, color = 'red') +
  scale_x_continuous(breaks = seq(0, .2, .025))

# Median
median(rw$chlorides)
```
There are several major outliers for chloride measurment, but around the mean and median the distribution is normal.


```{r Sulfur Dioxide, echo=FALSE}
ggplot(data = rw, aes(free.sulfur.dioxide)) +
  geom_histogram( color = 'red', binwidth = 2) +
  scale_x_continuous(breaks = seq(0, 75, 8)) +
  labs(x = 'Free Sulfur Dioxide (mg / dm^3)')

ggplot(data = subset(rw, total.sulfur.dioxide < quantile(total.sulfur.dioxide, .99)), 
       aes(total.sulfur.dioxide)) +
  geom_histogram(color = 'red', binwidth = 5) +
  scale_x_continuous(breaks = seq(0, 150, 15)) +
  labs(x = 'Total Sulfur Dioxide (mg / dm^3)')
```
Both measueres of sulfur dioxide are positively skewed. This makes sense because according to the description of the data "SO2 concentrations over 50 ppm [are] evident in the nose and taste of wine."   

```{r Sulfur above 50 ppm}
# Number of entries with sulfur ppm above 50.
dim(subset(rw, total.sulfur.dioxide >= 50))[1]
557 / 1599
```
Out of the 1599 entries there are 557 with SO2 above 50. This is about 34.8% of the entries.

```{r Residual Sugars, echo=FALSE}
ggplot(data = rw, 
       aes(residual.sugar)) +
  geom_histogram(color = 'red', binwidth = 0.2)

ggplot(data = subset(rw,residual.sugar < quantile(residual.sugar, .99)),
       aes(residual.sugar)) +
  geom_histogram(color = 'red', binwidth = 0.2) +
  scale_x_continuous(breaks = seq(0, 8, .4))

summary(rw$residual.sugar)
```
Residual sugars are normal with outliers.  
This seems to be a reoccurring pattern. Later it may be interesting to compare the tails with each other to see if there is a correlation between the extremes and good or bad wine.

```{r Density, echo=FALSE}
ggplot(data = rw, aes(density)) +
  geom_histogram(binwidth = .0005, color = 'red') +
  scale_x_continuous(breaks = seq(.985, 1.01, 0.0015))
```
Density is tightly distributed with a normal distribution showing no apparent skewness.

```{r Sulphates, echo=FALSE}
ggplot(data = rw, aes(sulphates)) +
  geom_histogram(binwidth = .03, color = 'red')

ggplot(data = subset(rw, sulphates < quantile(sulphates, .99)), 
       aes(sulphates)) +
  geom_histogram(binwidth = .03, color = 'red') +
  scale_x_continuous(breaks = seq(0, 1.25, .075))

summary(rw$sulphates)
```
Positively skewed with extreme positive outliers. Also, the entries are fairly spread out. According to the data file sulphates "can contribute to sulfur dioxide gas".

```{r Sulphates and sulfur dioxide, echo=FALSE}
plt1 <- ggplot(data = rw, aes(sulphates)) +
  geom_histogram(binwidth = .03, color = 'red')

plt2 <- ggplot(data = rw, aes(total.sulfur.dioxide)) +
  geom_histogram(color = 'red', binwidth = 5) +
  scale_x_continuous(breaks = seq(0, 300, 15)) +
  labs(x = 'Total Sulfur Dioxide (mg / dm^3)')

grid.arrange(plt1, plt2, ncol = 1)
```
Both do have similar tails, but the total sulfur dioxide is not as normal as the sulphates